---
title: 'Safepas: Mandal - Smolt movment'
author: "Kim Magnus Bærum"
date: ""
output: html_document
---

---
```{r, echo=FALSE}
# this code chunck is for setting global knitr options - does not show in document
library(knitr)
opts_chunk$set(tidy=TRUE, dpi=150)
```

**Date:** `r Sys.Date()`



## Data input 
It might be nice to work from an external database? So we always have data-backup, and always work with the latest versions..
```{r,warning=FALSE,message=FALSE,echo=TRUE,results='hide',eval=FALSE}
#setting directory
your_directory<-"C:/Users/kim.barum/Documents/Safepas"
setwd(your_directory)
library(moveHMM)
# Small grid (near intake)
#smolt_yaps2_SG<-read.csv("Smolt Hyd_SG_yap2_4.csv", header=T,sep=";",dec = "," ,na.strings = "")

# Large grid (rest of stream section)
#smolt_yaps2_LG<-read.csv("Smolt Hyd_LG_yap2_4.csv", header=T,sep=";",dec = "," ,na.strings = "")


smolt_yaps2_LG<-read.csv("SmoltSSIIM.SBG.Env.csv", header=T)

```

## Data wrangeling
Much more to be done here, but this represents a start
```{r,warning=FALSE,message=FALSE,echo=TRUE,results='hide',eval=FALSE}

# extracting some variables of interest 
data_LG<-smolt_yaps2_LG[, c("tag","fishx","fishy","u","v","w","k","eps","Discharge.sim")]
colnames(data_LG)[1] <- "ID" # for easy implementation in package 

model_data_LG<-prepData(data_LG,type="UTM",coordNames=c("fishx","fishy"))
head(model_data_LG)
summary(model_data_LG)

plot(model_data_LG,compact=T)

#seems best to remove fish with id: 4, 5 (only 6 points) 
model_data_LG<-model_data_LG[!(model_data_LG$ID=="4"| model_data_LG$ID=="5"),]


#standardize covariate values
#model_data_LG$v<-(model_data_LG$v-mean(model_data_LG$v))/sd(model_data_LG$v) 
#model_data_LG$k<-(model_data_LG$k-mean(model_data_LG$k))/sd(model_data_LG$k) 
#model_data_LG$eps<-(model_data_LG$eps-mean(model_data_LG$eps))/sd(model_data_LG$eps) 

# Making a datasett without observations that "stand out" as different, either with very many points, or very few points according to mean numbers of points pr fish
#model_data_LG_cleaned<-model_data_LG[!(model_data_LG$ID=="38"| model_data_LG$ID=="40"| model_data_LG$ID=="71"| model_data_LG$ID=="8"| #model_data_LG$ID=="35"| model_data_LG$ID=="37"| model_data_LG$ID=="50"| model_data_LG$ID=="53"| model_data_LG$ID=="85"| #model_data_LG$ID=="99"| model_data_LG$ID=="12"),]

#plot(model_data_LG_cleaned,compact=T)


```

## Data Analysis
The following analysis uses the fish tracks to evaluate state-transition probabilities in relation to environmental variables in the stream (or what ever variables one would want to include). For now I only included data from the large grid, as it seems most sensible to understand the variation in movement here (with the assumption that all fish that goes into the small-grid area are doomed anyway).


```{r,warning=FALSE,message=FALSE,echo=TRUE,results='hide',eval=FALSE}
# Model with all individuals
## initial parameters for gamma (step length) and von Mises distributions (turning angle), mimmicining restricted random walk (assuming 2 states: )
mu0_2<-c(2,3)# step mean (two parameters: one for each state)
sigma0_2<-c(2,2)# step SD
zeromass0_2<-c(0.05,0.05)# step zero-mass
stepPar0_2<-c(mu0_2,sigma0_2,zeromass0_2)
angleMean0_2<-c(0,0)# angle mean
kappa0_2<-c(4,0.5)# angle concentration
anglePar0_2<-c(angleMean0_2,kappa0_2)


## initial parameters for gamma and von Mises distributions (assuming 3 states)
mu0_3<-c(2,0.5,3)# step mean (3 parameters: one for each state)
sigma0_3<-c(1,0.5,3)# step SD
zeromass0_3<-c(0.005,0.05,0.05)# step zero-mass
stepPar0_3<-c(mu0_3,sigma0_3,zeromass0_3)
angleMean0_3<-c(0,0,pi)# angle mean
kappa0_3<-c(2,1,0.5)# angle concentration
anglePar0_3<-c(angleMean0_3,kappa0_3)


## initial parameters for gamma and von Mises distributions (assuming 4 states)
mu0_4<-c(2,0.5,3,1)# step mean (4 parameters: one for each state)
sigma0_4<-c(1,0.5,3,1)# step SD
zeromass0_4<-c(0.005,0.05,0.05,0.05)# step zero-mass
stepPar0_4<-c(mu0_4,sigma0_4,zeromass0_4)
angleMean0_4<-c(0,0,pi,pi)# angle mean
kappa0_4<-c(2,1,0.5,1)# angle concentration
anglePar0_4<-c(angleMean0_4,kappa0_4)

mu0_5<-c(2,0.5,3,1,1)# step mean (5 parameters: one for each state)
sigma0_5<-c(1,0.5,3,1,1)# step SD
zeromass0_5<-c(0.005,0.05,0.05,0.05,0.05)# step zero-mass
stepPar0_5<-c(mu0_5,sigma0_5,zeromass0_5)
angleMean0_5<-c(0,0,pi,pi,0)# angle mean
kappa0_5<-c(2,1,0.5,1,1)# angle concentration
anglePar0_5<-c(angleMean0_5,kappa0_5)

mu0_6<-c(2,0.5,3,1,1,10)# step mean (5 parameters: one for each state)
sigma0_6<-c(1,0.5,3,1,1,5)# step SD
zeromass0_6<-c(0.005,0.05,0.05,0.05,0.05,0.05)# step zero-mass
stepPar0_6<-c(mu0_6,sigma0_6,zeromass0_6)
angleMean0_6<-c(0,0,pi,pi,0,0)# angle mean
kappa0_6<-c(2,1,0.5,1,1,0.5)# angle concentration
anglePar0_6<-c(angleMean0_6,kappa0_6)


###
# Checking for numbers of states at a population level
m_2<-fitHMM(data=model_data_LG,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~1)
m_3<-fitHMM(data=model_data_LG,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~1)
m_4<-fitHMM(data=model_data_LG,nbStates=4,stepPar0=stepPar0_4,anglePar0=anglePar0_4,formula=~1)
m_5<-fitHMM(data=model_data_LG,nbStates=5,stepPar0=stepPar0_5,anglePar0=anglePar0_5,formula=~1)
m_6<-fitHMM(data=model_data_LG,nbStates=6,stepPar0=stepPar0_6,anglePar0=anglePar0_6,formula=~1)


AIC(m_2,m_3,m_4,m_5,m_6)

m_5v<-fitHMM(data=model_data_LG,nbStates=5,stepPar0=stepPar0_5,anglePar0=anglePar0_5,formula=~v)
m_5u<-fitHMM(data=model_data_LG,nbStates=5,stepPar0=stepPar0_5,anglePar0=anglePar0_5,formula=~u)
m_5k<-fitHMM(data=model_data_LG,nbStates=5,stepPar0=stepPar0_5,anglePar0=anglePar0_5,formula=~k)
m_5u+v<-fitHMM(data=model_data_LG,nbStates=5,stepPar0=stepPar0_5,anglePar0=anglePar0_5,formula=~u+v)
m_5u+v+k<-fitHMM(data=model_data_LG,nbStates=5,stepPar0=stepPar0_5,anglePar0=anglePar0_5,formula=~u+v+k)



## call to fitting function (this might take som while, the function runs mcmc-aproximations)
m<-fitHMM(data=model_data_LG,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~1)
m1<-fitHMM(data=model_data_LG,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~v)
m2<-fitHMM(data=model_data_LG,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~u)
m3<-fitHMM(data=model_data_LG,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~u+v)
m4<-fitHMM(data=model_data_LG,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~u*v)
m5<-fitHMM(data=model_data_LG,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~k)
m6<-fitHMM(data=model_data_LG,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~eps)

## Assessing models with AIC

AIC(m,m2,m3,m4,m5,m6)
  HMM.plot(move.HMM, xlim, breaks, by = NULL)
m_3<-fitHMM(data=model_data_LG,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~1)
m1_3<-fitHMM(data=model_data_LG,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~v)
m2_3<-fitHMM(data=model_data_LG,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~u)
m3_3<-fitHMM(data=model_data_LG,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~u+v)
m4_3<-fitHMM(data=model_data_LG,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~u*v)
m5_3<-fitHMM(data=model_data_LG,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~k)
m6_3<-fitHMM(data=model_data_LG,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~eps)

AIC(m,m2,m3,m4,m5,m6,m_3,m1_3,m2_3,m3_3,m4_3,m5_3,m6_3)

```

It seems that the individual variation is to big to be tackled in one go...In other words, it doesent seem valid to assume that all individuals share the same movment parameters, or behave accordingly to changes in the environmental parameters. 
In the next section I have done the same as above but per individual for an selection of 10 individuals from the full dataset.

```{r,warning=FALSE,message=FALSE,echo=TRUE,results='hide',eval=FALSE}
#Choose a random selection of 10 individuals
#sample(1:100,10)
# 27,59, 10,20,69,73,98,90,89,21
#Number 27
ind_27_data<-model_data_LG[model_data_LG$ID==27,]
plot(ind_27_data)# looks initially as the fish in general have 2 states, one with large turns and one with no turns, steps mi9gth also #differ
mu0_2<-c(2,2)# step mean (two parameters: one for each state)
sigma0_2<-c(2,2)# step SD
stepPar0_2<-c(mu0_2,sigma0_2)
angleMean0_2<-c(0,0.5)# angle mean
kappa0_2<-c(1,0.5)# angle concentration
anglePar0_2<-c(angleMean0_2,kappa0_2)

m_27<-fitHMM(data=ind_27_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~1)
m1_27<-fitHMM(data=ind_27_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~v)
m2_27<-fitHMM(data=ind_27_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~u)
m3_27<-fitHMM(data=ind_27_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~u+v)
m4_27<-fitHMM(data=ind_27_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~u*v)
m5_27<-fitHMM(data=ind_27_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~k)
m6_27<-fitHMM(data=ind_27_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~eps)

AIC(m_27,m1_27,m2_27,m3_27,m4_27,m5_27,m6_27)
plot(m4_27)
# state probabilities
sp <- stateProbs(m4_27)
ind_27_data$prob1<-sp[,1]
ind_27_data$prob2<-sp[,2]

ind_27_data$states <- viterbi(m4_27)

#Number 59
ind_59_data<-model_data_LG[model_data_LG$ID==59,]
plot(ind_59_data)# looks initially as the fish in general have 2 states, one with large turns and one with no turns, steps migth also #differ
mu0_2<-c(1,1)# step mean (two parameters: one for each state)
sigma0_2<-c(1,1)# step SD
zeromass0_2<-c(0.1,0.1)
stepPar0_2<-c(mu0_2,sigma0_2,zeromass0_2)
angleMean0_2<-c(0,0)# angle mean
kappa0_2<-c(0.5,1)# angle concentration
anglePar0_2<-c(angleMean0_2,kappa0_2)

m_59<-fitHMM(data=ind_59_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~1)
m1_59<-fitHMM(data=ind_59_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~v)
m2_59<-fitHMM(data=ind_59_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~u)
m3_59<-fitHMM(data=ind_59_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~u+v)
m4_59<-fitHMM(data=ind_59_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~u*v)
m5_59<-fitHMM(data=ind_59_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~k)
m6_59<-fitHMM(data=ind_59_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~eps)

AIC(m_59,m1_59,m2_59,m3_59,m4_59,m5_59,m6_59)

plot(m_59)
plotPR(m6_59)

# Will try with 3 states
## initial parameters for gamma and von Mises distributions (assuming 3 states)
mu0_3<-c(2,0.5,3)# step mean (3 parameters: one for each state)
sigma0_3<-c(1,0.5,3)# step SD
zeromass0_3<-c(0.05,0.05,0.05)# step zero-mass
stepPar0_3<-c(mu0_3,sigma0_3,zeromass0_3)
angleMean0_3<-c(pi,0,0)# angle mean
kappa0_3<-c(2,1,0.5)# angle concentration
anglePar0_3<-c(angleMean0_3,kappa0_3)

m_59_3<-fitHMM(data=ind_59_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~1)
m1_59_3<-fitHMM(data=ind_59_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~v)
m2_59_3<-fitHMM(data=ind_59_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~u)
m3_59_3<-fitHMM(data=ind_59_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~u+v)
m4_59_3<-fitHMM(data=ind_59_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~u*v)
m5_59_3<-fitHMM(data=ind_59_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~k)
m6_59_3<-fitHMM(data=ind_59_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~eps)
m7_59_3<-fitHMM(data=ind_59_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~u+v+k+eps)
m8_59_3<-fitHMM(data=ind_59_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~u+v+u*v)
m9_59_3<-fitHMM(data=ind_59_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~u+v+k)

AIC(m_59,m1_59,m2_59,m3_59,m4_59,m5_59,m6_59,m_59_3,m1_59_3,m2_59_3,m3_59_3,m4_59_3,m5_59_3,m6_59_3,m7_59_3,m9_59_3)
plot(m2_59_3)
plotPR(m2_59_3)
ind_59_data$states <- viterbi(m5_59_3)

# Number 10
ind_10_data<-model_data_LG[model_data_LG$ID==10,]
plot(ind_10_data)#

# Number 73
ind_73_data<-model_data_LG[model_data_LG$ID==73,]

plot(ind_73_data)

# Number 99
ind_99_data<-model_data_LG[model_data_LG$ID==99,]

plot(ind_99_data)


# Number 89
ind_89_data<-model_data_LG[model_data_LG$ID==89,]

plot(ind_89_data)

#two states
mu0_2<-c(1,1)# step mean (two parameters: one for each state)
sigma0_2<-c(1,1)# step SD
zeromass0_2<-c(0.05,0.05)
stepPar0_2<-c(mu0_2,sigma0_2)
angleMean0_2<-c(pi,0)# angle mean
kappa0_2<-c(1,1)# angle concentration
anglePar0_2<-c(angleMean0_2,kappa0_2)

m_89<-fitHMM(data=ind_89_data,nbStates=2,stepPar0=stepPar0_2,anglePar0=anglePar0_2,formula=~1)

#3 states

mu0_3<-c(2,0.5,3)# step mean (3 parameters: one for each state)
sigma0_3<-c(1,0.5,3)# step SD
zeromass0_3<-c(0.05,0.05,0.05)# step zero-mass
stepPar0_3<-c(mu0_3,sigma0_3)
angleMean0_3<-c(pi,0,0)# angle mean
kappa0_3<-c(2,1,0.5)# angle concentration
anglePar0_3<-c(angleMean0_3,kappa0_3)

m_89_3<-fitHMM(data=ind_89_data,nbStates=3,stepPar0=stepPar0_3,anglePar0=anglePar0_3,formula=~1)

# 4 states
mu0_4<-c(2,0.5,3,10)# step mean (3 parameters: one for each state)
sigma0_4<-c(1,0.5,3,5)# step SD
zeromass0_4<-c(0.05,0.05,0.05,0.005)# step zero-mass
stepPar0_4<-c(mu0_4,sigma0_4)
angleMean0_4<-c(pi,0,0,0)# angle mean
kappa0_4<-c(2,1,0.5,2)# angle concentration
anglePar0_4<-c(angleMean0_4,kappa0_4)

m_89_4<-fitHMM(data=ind_89_data,nbStates=4,stepPar0=stepPar0_4,anglePar0=anglePar0_4,formula=~1)

AIC(m_89,m_89_3,m_89_4)

plot(m_89_4)
m_89_4<-fitHMM(data=ind_89_data,nbStates=4,stepPar0=stepPar0_4,anglePar0=anglePar0_4,formula=~1)
m2_89_4<-fitHMM(data=ind_89_data,nbStates=4,stepPar0=stepPar0_4,anglePar0=anglePar0_4,formula=~u)
m3_89_4<-fitHMM(data=ind_89_data,nbStates=4,stepPar0=stepPar0_4,anglePar0=anglePar0_4,formula=~v)
m4_89_4<-fitHMM(data=ind_89_data,nbStates=4,stepPar0=stepPar0_4,anglePar0=anglePar0_4,formula=~k)
m5_89_4<-fitHMM(data=ind_89_data,nbStates=4,stepPar0=stepPar0_4,anglePar0=anglePar0_4,formula=~u+v)
m6_89_4<-fitHMM(data=ind_89_data,nbStates=4,stepPar0=stepPar0_4,anglePar0=anglePar0_4,formula=~u*v)
m7_89_4<-fitHMM(data=ind_89_data,nbStates=4,stepPar0=stepPar0_4,anglePar0=anglePar0_4,formula=~eps)
m8_89_4<-fitHMM(data=ind_89_data,nbStates=4,stepPar0=stepPar0_4,anglePar0=anglePar0_4,formula=~u+v+k+eps)

AIC(m_89_4, m2_89_4,m3_89_4,m4_89_4,m5_89_4,m6_89_4,m7_89_4,m8_89_4)

plot(simData(nbAnimals = 100,,obsPerAnimal=50,model=m2_89_4,states=TRUE))


```

## Including Plots


```{r, echo=FALSE}
library(sp)
#nb 27
utmcoor<-SpatialPoints(cbind(ind_27_data$x,ind_27_data$y), proj4string=CRS("+proj=utm +zone=32 +ellps=WGS84"))
loc1_geo<- spTransform(utmcoor, CRS("+proj=longlat +ellps=WGS84"))
ind_27_data$longlat_x<-loc1_geo$coords.x1
ind_27_data$longlat_y<-loc1_geo$coords.x2

# nb 59
utmcoor<-SpatialPoints(cbind(ind_59_data$x,ind_59_data$y), proj4string=CRS("+proj=utm +zone=32 +ellps=WGS84"))
loc1_geo<- spTransform(utmcoor, CRS("+proj=longlat +ellps=WGS84"))
ind_59_data$longlat_x<-loc1_geo$coords.x1
ind_59_data$longlat_y<-loc1_geo$coords.x2

#nb 73
utmcoor<-SpatialPoints(cbind(ind_73_data$x,ind_73_data$y), proj4string=CRS("+proj=utm +zone=32 +ellps=WGS84"))
loc1_geo<- spTransform(utmcoor, CRS("+proj=longlat +ellps=WGS84"))
ind_73_data$longlat_x<-loc1_geo$coords.x1
ind_73_data$longlat_y<-loc1_geo$coords.x2


library(ggmap)
mandal<-get_googlemap(center = c(lon = 7.5270, lat = 58.29627), zoom = 19,
              size = c(640, 640), scale = 2, format = "png8")

ind_59_data$states<-as.factor(ind_59_data$states)
ind_27_data$states<-as.factor(ind_27_data$states)
myColors <- c("orange", "blue", "green")
ggmap(mandal)+
  geom_point(aes(x = longlat_x, y = longlat_y,colour=states), data = ind_59_data)+scale_color_manual(values=myColors)  geom_line(aes(x = longlat_x, y = longlat_y), data = ind_59_data)+ 
  geom_point(aes(x = longlat_x, y = longlat_y,colour=states), data = ind_27_data)+
geom_point(aes(x = longlat_x, y = longlat_y), data = ind_73_data)



data2[data2$Discharge.sim==98,]

utmcoor<-SpatialPoints(cbind(data$fishx,data$fishy), proj4string=CRS("+proj=utm +zone=32 +ellps=WGS84"))
loc1_geo<- spTransform(utmcoor, CRS("+proj=longlat +ellps=WGS84"))
data$longlat_x<-loc1_geo$coords.x1
data$longlat_y<-loc1_geo$coords.x2
utmcoor<-SpatialPoints(cbind(data2$fishx,data2$fishy), proj4string=CRS("+proj=utm +zone=32 +ellps=WGS84"))
loc1_geo<- spTransform(utmcoor, CRS("+proj=longlat +ellps=WGS84"))
data2$longlat_x<-loc1_geo$coords.x1
data2$longlat_y<-loc1_geo$coords.x2
mandal<-get_googlemap(center = c(lon = 7.52750, lat = 58.29620), zoom = 18,
                      size = c(640, 640), scale = 2, format = "png8", maptype = "satellite")

data$tag2<-as.factor(data$tag)
data2$tag2<-as.factor(data2$tag)
ggmap(mandal)+
  geom_point(aes(x = longlat_x, y = longlat_y,colour=tag2), data = data2[data2$Discharge.sim==98,])+theme(legend.position="none")

ggmap(mandal)+
  geom_point(aes(x = longlat_x, y = longlat_y,colour=tag2), data = data[data$Discharge.sim==98,])+theme(legend.position="none")



```

```{r, echo=FALSE}
## Model movment based on changes in xy-coordinates

smolt_yaps2_LG

```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
